Fatalities, injuries, property and crop damage caused by weather events
=======================================================================

# Synopsis #


In this note we show that tornado, thunderstorm wind and heat are most
harmful with respect to population health in the United State. We also
show that flood, hurricane and tornado cause most property damage
while drought and flood cause most crop damage. we use U.S. National Oceanic and Atmospheric
Administration's (NOAA) storm database. This database tracks
characteristics of major storms and weather events in the United States, including when and where they occur.  It also estimates of any fatalities, injuries, and property damage.

# Data Processing #

In this section we present code we used for data processing.

## Load required libraries, set directories and fileahash ##

```{r, echo = TRUE}
require(data.table)
require(ggplot2)
require(filehash)
require(knitr)

if (!file.exists('data')) {dir.create('data')}
dbPath <- 'data/DB'
if (!file.exists(dbPath)) {dbCreate(dbPath)}
db <-  dbInit(dbPath)
```

## Download and read data ##


```{r, echo = TRUE}
download.storm.dt <- function() {
  fileURI <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
  fileBzPath <- 'data/repdata-data-StormData.csv.bz2'
  if (!file.exists(fileBzPath)) {
    download.file(fileURI, destfile=fileBzPath, method="curl")
  }
  storm.df <- read.csv(bzfile(fileBzPath))
  as.data.table(storm.df)
}


get.or.cache <- function(fun, svar) {
  if (dbExists(db, svar)) {
    #We check if svar is cached. If not we exectute function and cache
    #its value.

    print("Reading data.table form cache...")
    dbFetch(db, svar)
  } else {
    var <- fun()
    dbInsert(db, svar, var)
  }
}

storm.dt <- get.or.cache(download.storm.dt, "storm.dt")
str(storm.dt)


evtype <- unique(tolower(storm.dt[,EVTYPE]))
evtype



```


```{r}

strom.col.ordered <- function(cname) {
  sco <- storm.dt[,sum(get(cname)),by=EVTYPE]
  sco[order(sco$V1, decreasing=T)]
}

fatalities<-strom.col.ordered("FATALITIES")
cnum <- dim(fatalities)[1]
sum(fatalities[21:cnum]$V1)
fatalities[1:20]

injuries<-strom.col.ordered("INJURIES")
cnum <- dim(injuries)[1]
sum(injuries[21:cnum]$V1)
sum(injuries[1:20]$V1)
injuries[1:20]

```

```{r}
propexps <- storm.dt[,PROPDMGEXP]
cropexps <- storm.dt[,CROPDMGEXP]
propexps.unique <- sort(as.character(unique(propexps)))
cropexps.unique <- sort(as.character(unique(cropexps)))
propexps.unique
cropexps.unique

tonum <- function(ch) {
  ch <- as.character(ch)
  if (ch=='h' || ch =='H') {
    1
  } else if (ch == 'K'||ch == 'k') {
    3
  } else if (ch=='m'|| ch=='M') {
    6
  }
  else if (ch=='B') {
    9
  } else if (ch %in% c("",  "-", "?", "+")) {
    0
  } else {
    as.integer(ch)
  }
}

propexps.new <- get.or.cache(function() {sapply(propexps, tonum)}, "propexps.new")
cropexps.new <- get.or.cache(function() {sapply(cropexps, tonum)}, "cropexps.new")

storm.dt[, PropExp:=propexps.new]
storm.dt[, PropD:=PROPDMG*(10)^PropExp]
storm.dt[, CropExp:=cropexps.new]
storm.dt[, CropD:=CROPDMG*(10)^CropExp]

prop.dmg<-strom.col.ordered("PropD")
cnum <- dim(prop.dmg)[1]
sum(prop.dmg[21:cnum]$V1)
sum(prop.dmg[1:cnum]$V1)
prop.dmg[1:20]

crop.dmg<-strom.col.ordered("CropD")
cnum <- dim(crop.dmg)[1]
sum(crop.dmg[21:cnum]$V1)
sum(crop.dmg[1:cnum]$V1)
crop.dmg[1:20]

```
    
